## Manitou EF ICO Prescription

let's use the spatial tree list inventory for implementing the Individuals, Clumps, and Openings (ICO) method defined by Churchill et al. ([2016](https://scholarworks.umt.edu/ico/3)) to generate a cut-retention prescription. The `trees2ico` [package and workflow](https://github.com/georgewoolsey/trees2ico/tree/main) determines which trees to 'cut' or 'keep' based on stand-level targets for basal area and specific clump size proportions.

load the `trees2ico` [package](https://github.com/georgewoolsey/trees2ico/tree/main) which was developed specifically for this use-case

```{r, eval=FALSE}
## remotes helps us get packages hosted on github
install.packages("remotes")
## get trees2ico
remotes::install_github(repo = "georgewoolsey/trees2ico", upgrade = F)
```

### Define Targets

in order to define our basal area and tree clump group proportion targets, we need to know the current conditions. let's use some `trees2ico` helper functions to get the current conditions

first, we'll define our distance in meters to separate tree clump groupings. See [Churchill et al. (2016)](https://scholarworks.umt.edu/ico/3) Figure 4 (p.10) and [Matonis and Binkley (2018)](https://scholar.google.com/scholar?cluster=14613232085713826018&hl=en&as_sdt=0,6) who "calculated coverage of mosaic-meadows (percentage of stand > 6 m from overstory trees)" (p. 124)

[Churchill et al. (2016)](https://scholarworks.umt.edu/ico/3) define tree clumps:

>The definition of a tree clump is based on the average inter-tree distance at which mature/old trees of the dominant leave tree species have clearly interlocking crowns and form patches of contiguous canopy. This distance can vary from 15 to 22 feet (4.57 to 6.71 m), depending on site productivity of the stand. The ICO method requires that a single distance be used to define and identify clumps in the field. A default distance of 20 feet (6.1 m) is recommended for dry forests dominated by ponderosa pine and Douglas-fir. Trees are members of the same clump if they are within this distance of at least one other tree in the clump. (p.11)

```{r}
my_clump_dist_m <- 6
ico_clumps_temp <- trees2ico:::get_ico_clumps(treetops_sf, clump_dist_m = my_clump_dist_m) 
# huh?
names(ico_clumps_temp)
```

this gave us the tree list with clumps defined using a maximum distance between trees for determining tree clumps of `r scales::comma(my_clump_dist_m,accuracy=.1)` m which is based on the approximate crown radius of a typical, mature overstory tree. [Churchill et al. (2016)](https://scholarworks.umt.edu/ico/3) provide instructions for implementing the clump identification ([Plotkin et al. 2002](https://doi.org/10.1086/342823)) in ArcGIS:

>Use the Buffer tool (in the Proximity toolset within the Analysis toolbox) to create a buffer of distance $d/2$, one half the inter-tree distance, around each point. This quantity $d/2$ is meant to approximate the crown radius of a "typical" overstory tree. Set the Dissolve Type option to ALL, which dissolves overlapping buffers, creating a reduced set of spatially non-overlapping polygons stored as a multipart polygon feature...[Sanchez Meador et al. (2011)](https://scholar.google.com/scholar?cluster=15721451396848616769&oi=gsb&hl=en&as_sdt=0,6) provide some useful examples of how clump attributes can be summarized...The method described here can be modified to use measured or modeled crown radii for each tree in place of $d/2$
(p.36)

with respect to clump size groupings, [Churchill et al. (2016)](https://scholarworks.umt.edu/ico/3) note that:

>Proportions for clump sizes should be lumped into four or five bins for operational simplicity. We use 4 or 5 bins (Fig 5): individual trees, small clumps (2-4 trees), medium clumps (5-9 trees), and large clumps (10-20+ trees). Note that when instructed to leave a large clump (e.g. 10-20 trees), marking crews often have difficulty leaving the upper end of the size range (e.g. an 18, 19, or 20 tree clump). Thus adding a fifth bin for “super clumps” may be necessary (e.g. 15-20 trees or 20-25+ trees), especially if the upper size range of clumps is desired. (p. 12-13)

`trees2ico` implements these group sizes by default

```{r}
ico_clumps_temp$tree_list %>% 
  dplyr::pull(clump_n_trees_grp) %>% 
  # get factor levels in case there aren't any represented in our tree list
  levels()
```

use another `trees2ico` helper function to see the current distribution of trees by clump size in our data as well as the BA

```{r}
agg_stand_clumps_temp <- 
  ico_clumps_temp %>% 
  purrr::pluck("tree_list") %>%
  # get rid of columns we use as param
  dplyr::select( -dplyr::any_of(c(
    "hey_xxxxxxxxxx"
    , "stand_area_ha"
  ))) %>%
  trees2ico:::agg_stand_clumps(
    stand_area_ha = stand_sf$stand_area_ha
  )
```

current (pre-treatment) distribution of trees by clump size

```{r}
agg_stand_clumps_temp$props %>%
  dplyr::mutate(
    count=scales::comma(count,accuracy=1)
    , prop=scales::percent(prop,accuracy=0.1)
  ) %>% 
  kableExtra::kbl(
    caption = "current (pre-treatment) group size tree distribution"
    , col.names = c(
      "group size","# trees","% trees"
    )
  ) %>% 
  kableExtra::kable_styling()
```

what about BA?

```{r}
print(paste(
  "current stand BA:"
  , scales::comma(agg_stand_clumps_temp$stats$ba_m2_ha, accuracy = 0.1)
  , "m2/ha"
))
```

For setting the stand average density target, [Churchill et al. (2016)](https://scholarworks.umt.edu/ico/3) note:

>An average BA, TPA, or SDI target for the stand should be selected that is appropriate for the species, structure, site conditions, and management objectives. Expected mortality from prescribed fire should be factored in. Stand average targets can come from historical reference stands, plant association based stocking guides, density management tools, or a combination of both (see Franklin et al. (2013) for a full discussion of setting density targets). (p.11)

the `trees2ico` package requires a stand average basal area (BA) target in square meters per hectare since BA is commonly used in silvicultural managment and widely represented in the research on historical reference conditions.

starting from the current clump size proportions and basal area, we'll set targets for our ICO prescription based on historical reference conditions. [Churchill et al. (2016)](https://scholarworks.umt.edu/ico/3) suggest the following:

>In some cases, reference spatial information may already be available and summarized in a way that it can be directly incorporated into ICO prescriptions. Such data exist and have been published for areas in Arizona ([Abella and Denton 2009](https://doi.org/10.1139/X09-146), [Sánchez Meador et al. 2011](https://doi.org/10.1111/j.1526-100X.2010.00652.x)), the eastern Washington Cascades ([Churchill et al. 2013](https://doi.org/10.1016/j.foreco.2012.11.007)), the northern Rockies ([Larson et al. 2012](https://doi.org/10.1139/x2012-100)), and the Sierra Nevada ([Lydersen et al. 2013](https://doi.org/10.1016/j.foreco.2013.05.023)). Reference datasets for using ICO in other forest types, such as coastal Douglas-fir or Pacific silver fir, also exist ([Larson and Churchill 2008](https://doi.org/10.1139/X08-123)). (p.28)

relevant research for our study area and forest type includes:

* [Churchill et al. (2013)](https://doi.org/10.1016/j.foreco.2012.11.007)
* [Brown et al. (2015)](https://doi.org/10.1139/cjfr-2014-0387)
* [Ziegler et al. (2017)](https://doi.org/10.1016/j.foreco.2016.12.002)

define 3 target scenarios

```{r}
ico_targets_df <- 
  levels(ico_clumps_temp$tree_list$clump_n_trees_grp) %>% 
    dplyr::as_tibble() %>% 
    dplyr::rename(clump_n_trees_grp = 1) %>% 
    dplyr::mutate(
      clump_n_trees_grp = ordered(clump_n_trees_grp, levels =  levels(ico_clumps_temp$tree_list$clump_n_trees_grp))
    ) %>% 
    tidyr::uncount(3, .id = "scenario_id") %>% 
    dplyr::arrange(scenario_id, clump_n_trees_grp) %>% 
  dplyr::bind_cols(
    # scenarios
    dplyr::bind_rows(
      dplyr::tibble(
        scenario_desc = "Low Clumping (Churchill et al. 2013)" # all trees >15 cm dbh were mapped...The 15 cm cutoff was based on a commercial thinning minimum diameter threshold.
        , target_props = c(0.42, 0.37, 0.16, 0.05, 0.0)
        , target_ba_m2_ha = 15
      )
      , dplyr::tibble(
          scenario_desc = "Moderate Clumping (Ziegler et al. 2017)"
          , target_props = c(
            median(c(0.27,0.25,0.28,0.36,0.51,0.14,0.58))
            , median(c(0.38,0.36,0.37,0.35,0.24,0.28,0.42))
            , median(c(0.25,0.30,0.25,0.16,0.12,0.27,0.0))
            , 0.0 # median(c(0.11,0.10,0.06,0.08,0.12,0.15,0.0))
            , 0.0
          )
          , target_ba_m2_ha = median(c(14.5,19.7,10.6,9.5,9.0,17.6,7.5))
        ) %>% 
        dplyr::mutate(
          target_props = ifelse(
            dplyr::row_number() == 4
            , (1 - sum(target_props))
            , target_props
          )
        )
      , dplyr::tibble(
        scenario_desc = "High Clumping (Churchill et al. 2016 [Malheur])"
        , target_props = c(0.15, 0.25, 0.25, 0.1, 0.25)
        , target_ba_m2_ha = 21
      )
    )
  ) %>% 
  # set a constant BA target?
    # Brown et al. (2015): 25% q = 14.8, mean = 16.7; 75% q = 21.5
    # Ziegler et al. 2017: median(c(14.5,19.7,10.6,9.5,9.0,17.6,7.5))
  dplyr::mutate(
    target_ba_m2_ha = median(c(14.5,19.7,10.6,9.5,9.0,17.6,7.5))
    , scenario_desc_fact = scenario_desc %>% 
      stringr::str_extract("^.*?(?=\\()") %>% 
      stringr::str_squish() %>% 
      forcats::fct_inorder()
    # , huh = as.numeric(scenario_desc_fact)
  )
# dplyr::glimpse(ico_targets_df)
```

We set the same BA target of `scales::comma( max(ico_targets_df$target_ba_m2_ha), accuracy = 0.1 )` square meters per hectare for all of these scenarios. This target was based on post-treatment [Ziegler et al. (2017)](https://doi.org/10.1016/j.foreco.2016.12.002) outcomes on sites that had been treated across the Colorado Plateau and southern Rocky Mountains (matching our demonstration site) with the expressed treatment goals of restoring forest structural patterns and potential fire behavior akin to historic ranges.

what are the 3 target scenarios we defined?

```{r}
ico_targets_df %>% 
  dplyr::select(scenario_desc, clump_n_trees_grp, target_props) %>% 
  dplyr::mutate(target_props = scales::percent(target_props,accuracy=1)) %>% 
  tidyr::pivot_wider(names_from = clump_n_trees_grp, values_from = target_props) %>% 
  dplyr::mutate(scenario_desc = scenario_desc %>% stringr::str_replace("lumping \\(","lumping<br />\\(")) %>% 
  dplyr::rename(scenario=scenario_desc) %>% 
  kableExtra::kbl(
    caption = paste0(
      "ICO Scenarios: target proportion of trees in different sized clumps"
      , "<br />"
      , "Target BA for all: "
      , scales::comma( max(ico_targets_df$target_ba_m2_ha), accuracy = 0.1 )
      , " m<sup>2</sup> ha<sup>-1</sup>"
      , " (Brown et al. 2015)"
    )
    , escape = F
    # , digits = 2
  ) %>% 
  kableExtra::kable_styling() 
```

plot this

```{r}
pal_clump_grp <- 
  levels(ico_clumps_temp$tree_list$clump_n_trees_grp) %>% 
  length() %>% 
  harrypotter::hp(option = "Always") %>% 
  setNames(
    levels(ico_clumps_temp$tree_list$clump_n_trees_grp) %>% 
    as.character()
  )
# names(pal_clump_grp)
# scales::show_col(pal_clump_grp)

ico_targets_df %>% 
  dplyr::select(scenario_desc_fact, clump_n_trees_grp, target_props) %>% 
  dplyr::mutate(
    target_props_lab = scales::percent(ifelse(target_props>=0.08,target_props,NA), accuracy = 1)
    , scenario_desc_fact = forcats::fct_rev(scenario_desc_fact)
    , clump_n_trees_grp = forcats::fct_rev(clump_n_trees_grp)
    , huh = ifelse(
      clump_n_trees_grp %in% c(">15 trees","10-15 trees")
      , "white"
      , "black"
    )
  ) %>% 
ggplot2::ggplot(
  mapping = ggplot2::aes(x = target_props, y = scenario_desc_fact, fill=clump_n_trees_grp, group=clump_n_trees_grp)
) +
  ggplot2::geom_col(
    width = 0.7, alpha=0.95
  ) +
  ggplot2::geom_text(
    mapping = aes(
        label = target_props_lab
        , color = huh
        , fontface = "bold"
      )
    , position = ggplot2::position_stack(vjust = 0.5)
    # , color = "black"
    , size = 4
  ) +
  ggplot2::scale_fill_manual(values = pal_clump_grp) +
  ggplot2::scale_color_identity() +
  ggplot2::scale_x_continuous(labels = scales::percent_format()) +
  ggplot2::labs(
    fill = "clump\nsize"
    , y = ""
    , x = "target proportion of trees in different sized clumps"
    # , subtitle = ""
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "top"
    , legend.direction  = "horizontal"
    , legend.title = element_text(size=7)
    , axis.title.x = element_text(size=10, face = "bold")
    , axis.text.x = element_blank()
    , axis.ticks.x = element_blank()
    , axis.text.y = ggplot2::element_text(size = 9, face = "bold")
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(
      # nrow = 2, byrow = T, 
      reverse = T
      # , override.aes = list(alpha = 0.9)
    )
  )
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Get ICO Prescription

let's get the ICO prescrition for each of these target scenarios. We'll use the `trees2ico` package which uses a spatial tree list that includes DBH measurements in (cm) to generate a cut-retention prescription. It determines which trees to 'cut' or 'keep' based on stand-level targets for basal area and specific clump size proportions

```{r, results=F}
# outdir
ico_output_dir <- file.path(c2t_output_dir,"ico_prescriptions")
if(!dir.exists(ico_output_dir)) dir.create(ico_output_dir)
##############################################################
# map over the scenarios
##############################################################
trees2ico_ans_list <- 
  unique(ico_targets_df$scenario_id) %>% 
  purrr::set_names() %>% 
  purrr::map(function(x){
    # fnms
    fnm_tree_list <- file.path(ico_output_dir, paste0("tree_list_scenario_",x,".gpkg") )
    fnm_pre_cut_clumps <- file.path(ico_output_dir, paste0("pre_cut_clumps_scenario_",x,".gpkg") )
    fnm_post_cut_clumps <- file.path(ico_output_dir, paste0("post_cut_clumps_scenario_",x,".gpkg") )
    if(
      !file.exists( fnm_tree_list )
      || !file.exists( fnm_pre_cut_clumps )
      || !file.exists( fnm_post_cut_clumps )
    ){
      # trees2ico
      trees2ico_ans <- trees2ico::trees2ico(
        tree_list = treetops_sf %>% 
          dplyr::select( -dplyr::any_of(c(
            "hey_xxxxxxxxxx"
            , "stand_area_ha"
          )))
        , target_clump_props = ico_targets_df %>% 
          dplyr::filter(scenario_id==x) %>% 
          dplyr::arrange(clump_n_trees_grp) %>% 
          dplyr::pull(target_props)
        , target_ba_m2_ha = ico_targets_df %>% 
          dplyr::filter(scenario_id==x) %>% 
          dplyr::slice_max(order_by = target_ba_m2_ha, with_ties = F, na_rm = T) %>% 
          dplyr::pull(target_ba_m2_ha)
        , clump_dist_m = my_clump_dist_m
        , stand_area_ha = stand_sf$stand_area_ha[1]
      )
      # write
      trees2ico_ans$tree_list %>% sf::st_write( fnm_tree_list, append = F, quiet = T )
      trees2ico_ans$pre_cut_clumps %>% sf::st_write( fnm_pre_cut_clumps, append = F, quiet = T )
      trees2ico_ans$post_cut_clumps %>% sf::st_write( fnm_post_cut_clumps, append = F, quiet = T )
      # return
      return(trees2ico_ans)
    }else{
      return(list(
        "tree_list" = sf::st_read( fnm_tree_list, quiet =T ) %>% 
          dplyr::mutate(
            pre_clump_n_trees_grp = factor(
              pre_clump_n_trees_grp
              , ordered = T
              , levels = names(pal_clump_grp) # c("Individual", "2-4 trees", "5-9 trees", "10-15 trees", ">15 trees")
            )
            , post_clump_n_trees_grp = factor(
              post_clump_n_trees_grp
              , ordered = T
              , levels = names(pal_clump_grp) # c("Individual", "2-4 trees", "5-9 trees", "10-15 trees", ">15 trees")
            )
          )
        , "pre_cut_clumps" = sf::st_read( fnm_pre_cut_clumps, quiet =T ) %>% 
          dplyr::mutate(
            clump_n_trees_grp = factor(
              clump_n_trees_grp
              , ordered = T
              , levels = names(pal_clump_grp) # c("Individual", "2-4 trees", "5-9 trees", "10-15 trees", ">15 trees")
            )
          )
        , "post_cut_clumps" = sf::st_read( fnm_post_cut_clumps, quiet =T ) %>% 
          dplyr::mutate(
            clump_n_trees_grp = factor(
              clump_n_trees_grp
              , ordered = T
              , levels = names(pal_clump_grp) # c("Individual", "2-4 trees", "5-9 trees", "10-15 trees", ">15 trees")
            )
          )
      ))
    }
  })

```

what did we get back?

```{r}
trees2ico_ans_list[[1]] %>% names()
```

let's look at the `tree_list` object

```{r}
trees2ico_ans_list[[1]] %>% 
  purrr::pluck("tree_list") %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(
    treeID, tree_height_m, dbh_cm
    # cols that trees2ico::trees2ico says it created in the help file
    , is_keep_tree
    , pre_clump_id
    , pre_clump_n_trees
    , pre_clump_n_trees_grp
    , post_clump_id
    , post_clump_n_trees
    , post_clump_n_trees_grp
  ) %>% 
  dplyr::glimpse()
```

### Treatment Summary

compare the pre- and post-treatment proportion of trees in different sized clumps

```{r}
trees2ico_ans_trees_df <- 
  trees2ico_ans_list %>% 
  purrr::map(~purrr::pluck(.x,"tree_list")) %>% 
  purrr::list_rbind(names_to = "scenario_id") %>% 
  dplyr::mutate(
    scenario_id = as.numeric(scenario_id)
    ,cut_leave = ifelse(is_keep_tree, "leave", "cut") %>% ordered()
  ) %>% 
  dplyr::inner_join(
    ico_targets_df %>% 
      dplyr::distinct(scenario_id, scenario_desc_fact)
    , by = "scenario_id"
  ) %>% 
  sf::st_set_geometry("geom")
# count props
pre_post_temp <- 
  dplyr::bind_rows(
  # get pre-treatment
    dplyr::distinct(ico_targets_df, clump_n_trees_grp, scenario_desc_fact) %>% 
    dplyr::left_join(
      trees2ico_ans_trees_df %>% 
        sf::st_drop_geometry() %>%
        dplyr::group_by(scenario_desc_fact,pre_clump_n_trees_grp) %>% 
        dplyr::summarise(
          basal_area_m2_per_ha=sum(basal_area_m2)/stand_sf$stand_area_ha
          , trees = dplyr::n()
          , .groups = "drop"
        ) %>% 
        dplyr::ungroup() %>% 
        dplyr::rename(clump_n_trees_grp=pre_clump_n_trees_grp)
      , by = dplyr::join_by(clump_n_trees_grp, scenario_desc_fact)
    ) %>% 
    dplyr::group_by(scenario_desc_fact) %>% 
    dplyr::mutate(
      trees = dplyr::coalesce(trees,0)
      , pct_trees = trees/sum(trees,na.rm = T)
      , basal_area_m2_per_ha = sum(basal_area_m2_per_ha,na.rm = T)
    ) %>% 
    dplyr::ungroup() %>% 
    dplyr::mutate(status = "pre-treatment")
  # get post-treatment
  , dplyr::distinct(ico_targets_df, clump_n_trees_grp, scenario_desc_fact) %>% 
    dplyr::left_join(
      trees2ico_ans_trees_df %>% 
        sf::st_drop_geometry() %>%
        dplyr::filter(is_keep_tree) %>% 
        dplyr::group_by(scenario_desc_fact,post_clump_n_trees_grp) %>% 
        dplyr::summarise(
          basal_area_m2_per_ha=sum(basal_area_m2)/stand_sf$stand_area_ha
          , trees = dplyr::n()
          , .groups = "drop"
        ) %>% 
        dplyr::ungroup() %>% 
        dplyr::rename(clump_n_trees_grp=post_clump_n_trees_grp)
      , by = dplyr::join_by(clump_n_trees_grp, scenario_desc_fact)
    ) %>% 
    dplyr::group_by(scenario_desc_fact) %>% 
    dplyr::mutate(
      trees = dplyr::coalesce(trees,0)
      , pct_trees = trees/sum(trees,na.rm = T)
      , basal_area_m2_per_ha = sum(basal_area_m2_per_ha,na.rm = T)
    ) %>% 
    dplyr::ungroup() %>% 
    dplyr::mutate(status = "post-treatment")
) 
# table
pre_post_temp %>% 
  dplyr::select(-c(trees)) %>% 
  dplyr::mutate(
    pct_trees = scales::percent(pct_trees,accuracy=1)
    , basal_area_m2_per_ha = scales::comma(basal_area_m2_per_ha,accuracy=.1)
  ) %>% 
  tidyr::pivot_wider(names_from = clump_n_trees_grp, values_from = pct_trees) %>% 
  dplyr::rename(scenario=scenario_desc_fact) %>% 
  dplyr::arrange(scenario,desc(status)) %>% 
  dplyr::relocate(scenario,status) %>% 
  dplyr::rename(
    `BA m<sup>2</sup> ha<sup>-1</sup>` = basal_area_m2_per_ha
  ) %>% 
  kableExtra::kbl(
    caption = "ICO Scenarios: pre- and post-treatment proportion of trees in different sized clumps"
    , escape = F
    # , digits = 2
  ) %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::collapse_rows(columns = 1, valign = "top")
```

visualize that

```{r, fig.height=8}
pre_post_temp %>% 
dplyr::mutate(
    pct_trees_lab = scales::percent(ifelse(pct_trees>=0.05,pct_trees,NA), accuracy = 1)
    , clump_n_trees_grp = forcats::fct_rev(clump_n_trees_grp)
    , huh = ifelse(
      clump_n_trees_grp %in% c(">15 trees","10-15 trees")
      , "white"
      , "black"
    )
  ) %>% 
ggplot2::ggplot(
  mapping = ggplot2::aes(x = pct_trees, y = status, fill=clump_n_trees_grp, group=clump_n_trees_grp)
) +
  ggplot2::geom_col(
    width = 0.7, alpha=0.95
  ) +
  ggplot2::geom_text(
    mapping = aes(
        label = pct_trees_lab
        , color = huh
        , fontface = "bold"
      )
    , position = ggplot2::position_stack(vjust = 0.5)
    # , color = "black"
    , size = 4
  ) +
  # ggplot2::facet_grid(rows = dplyr::vars(scenario_desc_fact), switch = "y") +
  ggplot2::facet_wrap(facets = dplyr::vars(scenario_desc_fact), ncol = 1) +
  ggplot2::scale_fill_manual(values = pal_clump_grp) +
  ggplot2::scale_color_identity() +
  ggplot2::scale_x_continuous(labels = scales::percent_format()) +
  ggplot2::labs(
    fill = "clump\nsize"
    , y = ""
    , x = "target proportion of trees in different sized clumps"
    # , subtitle = ""
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "top"
    , legend.direction  = "horizontal"
    , legend.title = element_text(size=7)
    , axis.title.x = element_text(size=10, face = "bold")
    , axis.text.x = element_blank()
    , axis.ticks.x = element_blank()
    , axis.text.y = ggplot2::element_text(size = 9, face = "bold")
    # facet label
    , strip.placement = "outside"
    # , strip.background = ggplot2::element_blank() #element_rect(fill = "white", color = "black")
    , strip.text = ggplot2::element_text(size = 11, face = "bold", angle = 0, color = "black")
  ) +
  ggplot2::guides(
    fill = ggplot2::guide_legend(
      # nrow = 2, byrow = T, 
      reverse = T
      # , override.aes = list(alpha = 0.9)
    )
  )
```

#### Silvicultural metrics

Let's look at some common stand-level forestry metrics

```{r, message=F, results=F}
### stand-level summaries
silv_metrics_temp <-
  dplyr::bind_rows(
  treetops_sf %>%
    sf::st_drop_geometry() %>% 
    # dplyr::filter(dbh_cm >= ostory_dbh_cm) %>% 
    dplyr::ungroup() %>%
    dplyr::mutate(status = "pre-treatment") %>% 
    dplyr::group_by(status,stand_area_ha) %>%
    dplyr::summarise(
      n_trees = dplyr::n_distinct(treeID)
      , mean_dbh_cm = mean(dbh_cm, na.rm = T)
      , mean_tree_height_m = mean(tree_height_m, na.rm = T)
      , loreys_height_m = sum(basal_area_m2*tree_height_m, na.rm = T) / sum(basal_area_m2, na.rm = T)
      , basal_area_m2 = sum(basal_area_m2, na.rm = T)
      , sum_dbh_cm_sq = sum(dbh_cm^2, na.rm = T)
    ) %>%
    dplyr::ungroup() %>%
    dplyr::mutate(
      trees_per_ha = (n_trees/stand_area_ha)
      , basal_area_m2_per_ha = (basal_area_m2/stand_area_ha)
      , qmd_cm = sqrt(sum_dbh_cm_sq/n_trees)
    ) %>%
    dplyr::select(-c(sum_dbh_cm_sq)) %>% 
    dplyr::mutate(scenario_desc_fact = " ") %>% 
    dplyr::relocate(status,scenario_desc_fact)

  , trees2ico_ans_trees_df %>%
      sf::st_drop_geometry() %>% 
      dplyr::filter(is_keep_tree) %>%
      dplyr::ungroup() %>%
      dplyr::mutate(
        status = "post-treatment"
        , stand_area_ha = stand_sf$stand_area_ha[1]
      ) %>% 
      dplyr::group_by(status,scenario_desc_fact,stand_area_ha) %>%
      dplyr::summarise(
        n_trees = dplyr::n_distinct(treeID)
        , mean_dbh_cm = mean(dbh_cm, na.rm = T)
        , mean_tree_height_m = mean(tree_height_m, na.rm = T)
        , loreys_height_m = sum(basal_area_m2*tree_height_m, na.rm = T) / sum(basal_area_m2, na.rm = T)
        , basal_area_m2 = sum(basal_area_m2, na.rm = T)
        , sum_dbh_cm_sq = sum(dbh_cm^2, na.rm = T)
      ) %>%
      dplyr::group_by(scenario_desc_fact) %>%
      dplyr::mutate(
        trees_per_ha = (n_trees/stand_area_ha)
        , basal_area_m2_per_ha = (basal_area_m2/stand_area_ha)
        , qmd_cm = sqrt(sum_dbh_cm_sq/n_trees)
      ) %>%
      dplyr::select(-c(sum_dbh_cm_sq)) %>% 
      dplyr::relocate(status,scenario_desc_fact)
) %>% 
  dplyr::select(
    status,scenario_desc_fact
    , stand_area_ha
    , n_trees
    , mean_dbh_cm
    , qmd_cm
    , mean_tree_height_m
    , loreys_height_m
    , trees_per_ha
    , basal_area_m2_per_ha
  )
```

```{r}
# table
silv_metrics_temp %>% 
  dplyr::mutate(
    dplyr::across(
      .cols = c(stand_area_ha) 
      , .fns = ~ scales::comma(.x, accuracy = .1)
    )
    , dplyr::across(
      .cols = c(n_trees) 
      , .fns = ~ scales::comma(.x, accuracy = 1)
    )
  ) %>% 
  kableExtra::kbl(
    digits = 1
    , escape = F
    , caption = "ICO Scenarios: pre- and post-treatment silvicultral metrics"
    , col.names = c(
      "status","scenario"
      , "area (ha)"
      , "trees"
      , "mean<br>DBH (cm)"
      , "QMD (cm)"
      , "mean<br>Ht. (m)"
      , "Loreys<br>Ht. (m)"
      , "TPH"
      , "BA<br>m<sup>2</sup> ha<sup>-1</sup>"
      )
  ) %>% 
  kableExtra::kable_styling(font_size = 12) %>% 
  kableExtra::collapse_rows(columns = 1, valign = "top")
```

visualize that

```{r, fig.height=11.3, fig.width=8}
silv_metrics_temp %>% 
  dplyr::filter(status == "pre-treatment") %>% 
  dplyr::select(-scenario_desc_fact) %>% 
  tidyr::crossing(trees2ico_ans_trees_df %>% dplyr::distinct(scenario_desc_fact)) %>% 
  dplyr::mutate(scenario_desc_fact = as.character(scenario_desc_fact)) %>% 
  dplyr::relocate(status,scenario_desc_fact) %>% 
  dplyr::bind_rows(
    silv_metrics_temp %>% dplyr::filter(status == "post-treatment")
  ) %>% 
  dplyr::mutate(
    scenario_desc_fact = forcats::fct_inorder(scenario_desc_fact)
    , status = ordered(status) %>% forcats::fct_rev()
  ) %>% 
  dplyr::select(-c(stand_area_ha)) %>% 
  tidyr::pivot_longer(cols = -c(scenario_desc_fact,status)) %>% 
  dplyr::group_by(scenario_desc_fact,name) %>% 
  dplyr::arrange(scenario_desc_fact,name,status) %>% 
  dplyr::mutate(pct_diff = (value-dplyr::lag(value))/dplyr::lag(value)) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    lab = paste0(
      dplyr::case_when(
        name == "n_trees" ~ scales::comma(value,accuracy=1)
        , T ~ scales::comma(value,accuracy=0.1)
      )
      , dplyr::case_when(
        is.na(pct_diff) ~ ""
        , T ~ paste0(
          "\n"
          , ifelse(pct_diff<0,"-","+")
          ,scales::percent(abs(pct_diff),accuracy=0.1)
        )
      )
    )
    , metric = factor(
      name
      , ordered = T
      , levels = c(
          "n_trees"
          ,"mean_dbh_cm"
          ,"qmd_cm"
          ,"mean_tree_height_m"
          ,"loreys_height_m"
          ,"trees_per_ha"
          ,"basal_area_m2_per_ha"
        )
      , labels = c(
          latex2exp::TeX("Number of Trees", output = "character")
          , latex2exp::TeX("Mean DBH (cm)", output = "character")
          , latex2exp::TeX("QMD (cm)", output = "character")
          , latex2exp::TeX("Mean Tree Ht (m)", output = "character")
          , latex2exp::TeX("Lorey's Mean Ht (m)", output = "character")
          , latex2exp::TeX("Trees $ha^{-1}$")
          , latex2exp::TeX("BA $m^{2} \\cdot ha^{-1}$")
        )
    )
    , scenario_desc_fact = as.character(scenario_desc_fact) %>% stringr::str_replace_all(" ", " ~ ") %>% forcats::fct_inorder()
  ) %>% 
  ggplot2::ggplot(
    mapping = ggplot2::aes(
      x = status
      , y = value
      , label = lab
      , group = 1
    )
  ) +
  ggplot2::geom_line(key_glyph = "point", alpha = 0.7, color = "gray", lwd = 1.1) +
  ggplot2::geom_col(mapping = ggplot2::aes(fill = status), alpha = 1, width = 0.4) +
  ggplot2::scale_fill_viridis_d(option = "cividis") +
  ggplot2::geom_text(
    vjust = -0.25, size = 2.5
  ) +
  ggplot2::facet_grid(
    rows = dplyr::vars(metric)
    , cols = dplyr::vars(scenario_desc_fact)
    , scales = "free_y"
    , labeller = ggplot2::label_parsed
    , switch = "y"
    , axes = "all_x"
  ) +
  ggplot2::scale_y_continuous(labels = scales::comma, expand = ggplot2::expansion(mult = c(0,.3)), breaks = NULL) +
  ggplot2::labs(
    x = "", y = ""
    , subtitle = ""
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , axis.text.x = ggplot2::element_text(size = 8, color = "black")
    , strip.text.x = ggplot2::element_text(size = 11, color = "black", face = "bold")
    , strip.text.y = ggplot2::element_text(size = 9, color = "black")
    , panel.grid = ggplot2::element_blank()
  ) 
```

height and diameter distributions

```{r, fig.height = 8}
trees2ico_ans_list %>% 
  purrr::map(
    \(x)
    x %>% 
    purrr::pluck("tree_list") %>% 
    sf::st_drop_geometry() %>% 
    dplyr::select(tree_height_m, dbh_cm) %>% 
    tidyr::pivot_longer(dplyr::everything()) %>% 
    dplyr::mutate(status = "pre-treatment") %>% 
    dplyr::bind_rows(
      x %>% 
        purrr::pluck("tree_list") %>% 
        sf::st_drop_geometry() %>% 
        dplyr::filter(is_keep_tree) %>% 
        dplyr::select(tree_height_m, dbh_cm) %>% 
        tidyr::pivot_longer(dplyr::everything()) %>% 
        dplyr::mutate(status = "post-treatment")
    )
  ) %>% 
  purrr::list_rbind(names_to = "scenario_id") %>% 
  dplyr::mutate(scenario_id = as.numeric(scenario_id)) %>% 
  dplyr::inner_join(
    ico_targets_df %>% 
      dplyr::distinct(scenario_id, scenario_desc_fact)
    , by = "scenario_id"
  ) %>% 
  dplyr::mutate(
    status = ordered(status) %>% forcats::fct_rev()
    , name = dplyr::recode_values(
      name
      , "tree_height_m" ~ "Tree Height (m)"
      , "dbh_cm" ~ "DBH (cm)"
    )
  ) %>%
  ggplot2::ggplot(mapping = ggplot2::aes(x=value,fill = status,color = status)) +
    ggplot2::geom_density(mapping = ggplot2::aes(y=ggplot2::after_stat(scaled)), alpha = 0.7, color=NA) +
    # ggplot2::geom_histogram(alpha = 0.7) +
    ggplot2::facet_grid(
      cols = dplyr::vars(name)
      , rows = dplyr::vars(scenario_desc_fact)
      , switch = "y"
      , axes = "all_x"
      , scales = "free"
    ) +
    ggplot2::scale_y_continuous(NULL,breaks=NULL) +
    ggplot2::scale_x_continuous(limits = c(0,NA), labels = scales::comma) +
    ggplot2::scale_fill_viridis_d(option = "cividis") +
    ggplot2::scale_color_viridis_d(option = "cividis") +
    ggplot2::labs(
      fill="",x=""
      , subtitle = "comparison of height and diameter distributions\npre- versus post-treatment"
    ) +
    ggplot2::theme_light() +
    ggplot2::theme(
      legend.position = "top"
      , strip.text = ggplot2::element_text(size = 10, color = "black", face = "bold")
      , axis.text.x = ggplot2::element_text(size = 9)
      , axis.text.y = ggplot2::element_blank()
      , axis.ticks.y = ggplot2::element_blank()
    )
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Treatment Spatial Pattern

let's check out the spatial pattern resulting from the different treatment scenarios

make a function to get the distance between trees as a raster and the openings as vectors

```{r}
# create function to pass a return from get_clump_summary() and get a distance raster
get_clump_dist_rast <- function(clumps_sf, stand_sf, clump_dist_m){
  # rasterize the clump polygons and then calculate distance between clumps as raster
  dist_rast <- 
    terra::rasterize(
        x = clumps_sf %>% terra::vect()
        , y = clumps_sf %>% 
          terra::vect() %>% 
          terra::rast(res = 0.2)
      ) %>% 
      terra::distance() %>% 
      # crop it to stand extent
      terra::crop(
        stand_sf %>% 
          terra::vect()
        , mask = T
      )
  ######### part 2
  # now create openings vector data
  openings_vect <- 
    dist_rast %>% 
      terra::classify(rcl = c(clump_dist_m/2,Inf), others = NA, include.lowest = T) %>% 
      terra::as.polygons(na.rm = T) %>% 
      sf::st_as_sf() %>% 
      sf::st_cast("POLYGON") %>% 
      dplyr::mutate(layer = dplyr::row_number()) %>% 
      dplyr::mutate(
        openining_area_m2 = sf::st_area(geometry) %>% as.numeric()
        , clump_dist_m = clump_dist_m
      )
  
  # return
  return(list(dist_rast = dist_rast, openings_vect = openings_vect))
}
# do it with the pre-treatment data
pre_get_clump_dist_rast_ans <- get_clump_dist_rast(clumps_sf = trees2ico_ans_list[[1]]$pre_cut_clumps, stand_sf = stand_sf, clump_dist_m = my_clump_dist_m)
# pre_get_clump_dist_rast_ans$dist_rast %>% terra::plot()
# pre_get_clump_dist_rast_ans$openings_vect %>% terra::plot()
```

now we'll make a function to plot the spatial pattern roughly following [Churchill et al. (2013)](https://doi.org/10.1016/j.foreco.2012.11.007) Figure 2

```{r, results=F}
# scales::show_col(pal_clump_grp)
# get_the_plts 
get_the_plts_fn <- function(dist_rast, openings_vect, stand_sf, clump_polys, ttops, my_title = "", openining_area_m2_max = NA, dist_max = NA) {
  # get the area spacing
  if(is.na(openining_area_m2_max)){
    openining_area_m2_max <- 
      openings_vect %>% 
      dplyr::pull(openining_area_m2) %>% 
      max() %>% 
      ceiling()
  }
  if(is.na(dist_max)){
    dist_max <- 
      dist_rast %>% 
      terra::minmax() %>% 
      .[2] %>% 
      ceiling()
  }
  plt_stand <- 
    ggplot2::ggplot() +
      ggplot2::geom_sf(data = stand_sf %>% dplyr::slice(1), color = "black", fill = NA, lwd = 1.5) +
      ggplot2::scale_x_continuous(expand = c(0, 0)) +
      ggplot2::scale_y_continuous(expand = c(0, 0)) +
      ggplot2::labs(
        x = ""
        , y = ""
        , subtitle = my_title
      ) +
      ggplot2::theme_void() +
      ggplot2::theme(
        legend.position = "bottom" # c(0.5,1)
        , legend.direction = "horizontal"
        , legend.margin = ggplot2::margin(0,0,0,0)
        , legend.title= ggplot2::element_text(size=8, face = "bold")
        , legend.text= ggplot2::element_text(size = 7)
        , legend.key = ggplot2::element_rect(fill = NA, color = NA)
        , plot.subtitle = ggplot2::element_text(size = 10, hjust = 0.5, face = "bold")
      )
##################################################
# plot distance raster and trees
##################################################
plt_dist_rast1 <-
  # ggplot2::ggplot() + 
  plt_stand +
    # distance
    ggplot2::geom_tile(
      data = dist_rast %>% 
        # terra::aggregate(fact = 2, fun = "max", cores = 4) %>% 
        # terra::crop(
        #   stand_boundaries %>% 
        #     dplyr::filter(poly_name==x)
        #     # dplyr::filter(poly_name=="west")
        #   , mask = T
        # ) %>% 
        terra::as.data.frame(xy = T) %>% 
        dplyr::rename(f=3)
      , mapping = ggplot2::aes(x=x, y=y, fill = f)
    ) +
    harrypotter::scale_fill_hp(
      "mischief"
      , na.value = "transparent"
      , name = "distance to\nnearest tree (m)"
      , limits = c(0,dist_max)
    ) +
    # openings
    ggplot2::geom_sf(
      data = openings_vect
      , mapping = ggplot2::aes(color = openining_area_m2), fill = NA
    ) +
    ggplot2::scale_color_gradient(
      low = "gray77", high = "gray11"
      , labels = scales::comma_format(accuracy = 1)
      , name = "opening\narea (m²)" # latex2exp::TeX("opening\narea ($\\m^2$)")
      , limits = c(0,openining_area_m2_max)
    ) +
    # clumps
    ggnewscale::new_scale_fill() +
    ggplot2::geom_sf(
      data = clump_polys
      , mapping = ggplot2::aes(fill = clump_n_trees_grp)
      , color = NA
    ) +
    # harrypotter::scale_fill_hp_d("always", name = "clump size", drop = F) +
    ggplot2::scale_fill_manual(
      values = pal_clump_grp, name = "clump size", drop = F
      , guide = ggplot2::guide_legend(nrow = 2, byrow = T)
    ) +
    # tree points
    ggplot2::geom_sf(data = ttops, color = "gray88", shape = ".") +
    # ggplot2::labs(subtitle = paste0(nm, ": pre-treatment")) +
    ggplot2::theme_void() +
    ggplot2::theme(
      legend.position = "bottom"
      , legend.direction = "horizontal"
      , plot.subtitle = ggplot2::element_text(size = 10, hjust = 0.5, face = "bold")
      , legend.title= ggplot2::element_text(size=8, face = "bold")
      , legend.text= ggplot2::element_text(size = 7)
    )
##################################################
# highlight the openings
##################################################
  plt_open1 <-
    plt_stand +
      # clumps
      ggplot2::geom_sf(
        data = clump_polys
        , mapping = ggplot2::aes(fill = clump_n_trees_grp)
        , color = NA
      ) +
      # harrypotter::scale_fill_hp_d("always", name = "clump size", drop = F) + 
      ggplot2::scale_fill_manual(
        values = pal_clump_grp, name = "clump size", drop = F
        , guide = ggplot2::guide_legend(nrow = 2, byrow = T)
      ) +
      # openings
      ggnewscale::new_scale_fill() +
      ggplot2::geom_sf(
        data = openings_vect
        , mapping = ggplot2::aes(fill = openining_area_m2), color = NA
      ) +
      ggplot2::scale_fill_gradient(
        low = "gray77", high = "gray11"
        , labels = scales::comma_format(accuracy = 1)
        , name = "opening\narea (m²)" # latex2exp::TeX("opening\narea ($\\m^2$)")
        , limits = c(0,openining_area_m2_max)
      ) +
      # tree points
      ggplot2::geom_sf(data = ttops, color = "gray88", shape = ".") +
      # ggplot2::labs(subtitle = paste0(nm, ": pre-treatment")) +
      ggplot2::theme_void() +
      ggplot2::theme(
          legend.position = "bottom"
          , legend.direction = "horizontal"
          , plot.subtitle = ggplot2::element_text(size = 10, hjust = 0.5, face = "bold")
          , legend.title= ggplot2::element_text(size=8, face = "bold")
          , legend.text= ggplot2::element_text(size = 7)
        )
  
  return(list(
    plt_dist_rast=plt_dist_rast1
    , plt_open=plt_open1
  ))
}
```

get all the figures for the treatment scenarios and pre-treatment

```{r, results=F}
get_clump_dist_rast_ans <- 
  trees2ico_ans_list %>% 
  purrr::map(function(x){
    get_clump_dist_rast_ans <- 
      get_clump_dist_rast(clumps_sf = x$post_cut_clumps, stand_sf = stand_sf, clump_dist_m = my_clump_dist_m)
    return(get_clump_dist_rast_ans)
  })
# get max opening and dist
dist_max_temp <- 
  get_clump_dist_rast_ans %>% 
  purrr::map(
    \(x)
    x %>% 
      purrr::pluck("dist_rast") %>% 
      terra::minmax() %>% 
      .[2] %>% 
      ceiling()
  ) %>% 
  unlist() %>% 
  max(na.rm = T)
openining_area_m2_max_temp <- 
  get_clump_dist_rast_ans %>% 
  purrr::map(
    \(x)
    x %>% 
      purrr::pluck("openings_vect") %>% 
      dplyr::pull(openining_area_m2) %>% 
      max() %>% 
      ceiling()
  ) %>% 
  unlist() %>% 
  max(na.rm = T)
# plts
plt_post <- 
  trees2ico_ans_list %>% 
  purrr::imap(function(x,nm){
    # get_the_plts_fn
      plts <- get_the_plts_fn(
        dist_rast = get_clump_dist_rast_ans[[nm]]$dist_rast
        , openings_vect = get_clump_dist_rast_ans[[nm]]$openings_vect
        , stand_sf = stand_sf
        , clump_polys = x$post_cut_clumps
        , ttops = x$tree_list %>% dplyr::filter(is_keep_tree)
        , my_title = ico_targets_df %>% dplyr::filter(as.character(scenario_id)==nm) %>% dplyr::slice(1) %>% dplyr::pull(scenario_desc_fact) %>% as.character()
        , openining_area_m2_max = openining_area_m2_max_temp, dist_max = dist_max_temp
      )    
      return(plts)
  })
# plt_post[[1]]$plt_dist_rast
# get_the_plts_fn
plt_pre <- get_the_plts_fn(
  dist_rast = pre_get_clump_dist_rast_ans$dist_rast
  , openings_vect = pre_get_clump_dist_rast_ans$openings_vect
  , stand_sf = stand_sf
  , clump_polys = trees2ico_ans_list[[1]]$pre_cut_clumps
  , ttops = treetops_sf
  , my_title = "Pre-Treatment"
  , openining_area_m2_max = openining_area_m2_max_temp, dist_max = dist_max_temp
)

```

check those plots out

the distance raster

```{r,eval=T, fig.height=10.1, fig.width=9}
plt_post_temp <- 
  plt_post %>% 
  purrr::map(
    \(x)
    x$plt_dist_rast + 
      ggplot2::theme(legend.position = "none") +
      ggplot2::guides(color = "none", fill = "none", shape = "none")
  )

patchwork::wrap_plots(
  patchwork::wrap_plots(
    append(
      plt_post_temp
      , list(
        plt_pre$plt_dist_rast
      )
      , after = 0
    )
    , ncol = 2
  )
  , patchwork::guide_area()
  , ncol = 1
  , heights = c(1,0.02)
) +
patchwork::plot_layout(guides = "collect") +
patchwork::plot_annotation(
  theme = ggplot2::theme(
    legend.position = "bottom"
    , legend.margin = ggplot2::margin(b = 11, unit = "pt")
    , legend.box.margin = ggplot2::margin(b = 11, unit = "pt")
    # , plot.title = ggplot2::element_text(margin = ggplot2::margin(t = -5, b = 5, unit = "pt"))
  )
)

```

```{r,eval=T, include=FALSE, eval=TRUE}
ggplot2::ggsave(
  "../data/ico_treatment_dist_rast.jpg"
  , dpi = "print"
  , height = 10.1, width = 9
)
```

the openings

```{r,eval=T, fig.height=10.1, fig.width=9}
plt_post_temp <- 
  plt_post %>% 
  purrr::map(
    \(x)
    x$plt_open + 
      ggplot2::theme(legend.position = "none") +
      ggplot2::guides(color = "none", fill = "none", shape = "none")
  )
# plt_post_temp[[1]]
# plt_pre$plt_open
patchwork::wrap_plots(
  patchwork::wrap_plots(
    append(
      plt_post_temp
      , list(
        plt_pre$plt_open
      )
      , after = 0
    )
    , ncol = 2
  )
  , patchwork::guide_area()
  , ncol = 1
  , heights = c(1,0.02)
) +
patchwork::plot_layout(guides = "collect") +
patchwork::plot_annotation(
  theme = ggplot2::theme(
    legend.position = "bottom"
    , legend.margin = ggplot2::margin(b = 11, unit = "pt")
    , legend.box.margin = ggplot2::margin(b = 11, unit = "pt")
    # , plot.title = ggplot2::element_text(margin = ggplot2::margin(t = -5, b = 5, unit = "pt"))
  )
)

```

```{r,eval=T, include=FALSE, eval=TRUE}
ggplot2::ggsave(
  "../data/ico_treatment_openings.jpg"
  , dpi = "print"
  , height = 10.1, width = 9
)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Treatment Cut/Leave Trees

let's check out which trees were classified as 'cut' and which trees to 'leave' in each treatment scenario

remember that [sweet height contour plot](#pre_str) we made using pre-treatment trees and their heights? let's use that as a background...

```{r, fig.height = 10.1, fig.width = 9}
# plts
plt_cutleave <- 
  trees2ico_ans_list %>% 
  purrr::imap(function(x,nm){
    plt <- 
      plt_ht_contour +
      ggplot2::geom_sf(
        data = x$tree_list %>% 
          dplyr::mutate(cut_leave = ifelse(is_keep_tree, "leave", "cut") %>% ordered()) %>% 
          dplyr::arrange(cut_leave)
        , mapping = ggplot2::aes(color = cut_leave)
        , size = 1.1, alpha = 0.7
      ) +
      ggplot2::scale_color_manual(
        values = c("gray77","navy")
        , name = ""
      ) +
      ggplot2::labs(
        subtitle = ico_targets_df %>% dplyr::filter(as.character(scenario_id)==nm) %>% dplyr::slice(1) %>% dplyr::pull(scenario_desc_fact) %>% as.character()
      ) +
      ggplot2::guides(
        color = ggplot2::guide_legend(override.aes = list(shape = 15, size = 6, alpha = 1))
      )
    return(plt)
  })
# plt_cutleave[[1]]

# patch that
patchwork::wrap_plots(
  patchwork::guide_area()
  # , patchwork::plot_spacer()
  , patchwork::wrap_plots(
    append(
      plt_cutleave
      , list(plt_ht_contour_trees + ggplot2::labs(subtitle = "Pre-Treatment Tree Heights") )
      , after = 0
    )
    , ncol = 2
  )
  , ncol = 1
  , heights = c(0.01, 1)
) +
patchwork::plot_layout(guides = "collect") +
patchwork::plot_annotation(
  theme = ggplot2::theme(
    legend.position = "top"
    , legend.margin = ggplot2::margin(b = 0, unit = "pt")
    , legend.box.margin = ggplot2::margin(b = 0, unit = "pt")
    , plot.title = ggplot2::element_text(margin = ggplot2::margin(t = -5, b = 5, unit = "pt"))
  )
)
```

```{r, include=FALSE, eval=TRUE}
ggplot2::ggsave(
  "../data/ico_treatment_cutleave.jpg"
  , dpi = "print"
  , height = 10.1, width = 9
)
```

neat.

let's look at the distribution of heights by cut/leave tree group

```{r, fig.height=9}
trees2ico_ans_trees_df %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(scenario_desc_fact, cut_leave, tree_height_m, dbh_cm) %>% 
  tidyr::pivot_longer(cols = -c(scenario_desc_fact, cut_leave)) %>% 
  dplyr::mutate(
    name = dplyr::recode_values(
      name
      , "tree_height_m" ~ "Tree Height (m)"
      , "dbh_cm" ~ "DBH (cm)"
    )
  ) %>%
  ggplot2::ggplot(mapping = ggplot2::aes(x=value,fill = cut_leave,color = cut_leave)) +
    ggplot2::geom_density(mapping = ggplot2::aes(y=ggplot2::after_stat(scaled)), alpha = 0.7, color=NA) +
    # ggplot2::geom_histogram(alpha = 0.7) +
    ggplot2::facet_grid(
      cols = dplyr::vars(name)
      , rows = dplyr::vars(scenario_desc_fact)
      , switch = "y"
      , axes = "all_x"
      , scales = "free"
    ) +
    ggplot2::scale_y_continuous(NULL,breaks=NULL) +
    ggplot2::scale_x_continuous(limits = c(0,NA), labels = scales::comma) +
    ggplot2::scale_color_manual(
      values = c("gray77","navy")
      , name = ""
    ) +
    ggplot2::scale_fill_manual(
      values = c("gray77","navy")
      , name = ""
    ) +
    ggplot2::labs(
      fill="",x=""
      , subtitle = "comparison of height and diameter distributions\ncut versus leave trees"
    ) +
    ggplot2::theme_light() +
    ggplot2::theme(
      legend.position = "top"
      , strip.text = ggplot2::element_text(size = 10, color = "black", face = "bold")
      , axis.text.x = ggplot2::element_text(size = 9)
      , axis.text.y = ggplot2::element_blank()
      , axis.ticks.y = ggplot2::element_blank()
    )
```

another way to see it

```{r, fig.height=8.5}
trees2ico_ans_trees_df %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(scenario_desc_fact, cut_leave, tree_height_m, dbh_cm) %>% 
  tidyr::pivot_longer(cols = -c(scenario_desc_fact, cut_leave)) %>%
  dplyr::mutate(
    name = dplyr::recode_values(
      name
      , "tree_height_m" ~ "Tree Height (m)"
      , "dbh_cm" ~ "DBH (cm)"
    )
  ) %>%
  # dplyr::glimpse()
  ggplot2::ggplot(
    mapping = ggplot2::aes(y = value, x = cut_leave)
  ) +
  ggplot2::geom_boxplot(
    mapping = ggplot2::aes(fill = cut_leave)
    , width = 0.5
    , alpha = 0.88
    , color = "black"
    , outliers = F
  ) +
  # ggplot2::geom_text(
  #   mapping = ggplot2::aes(label = lab)
  #   , color = "black", size = 2.5, vjust = -0.2
  # ) +
  ggplot2::scale_fill_manual(
    values = c("gray77","navy")
    , name = ""
  ) +
  ggplot2::scale_x_discrete(drop = F) + 
  ggplot2::scale_y_continuous(
    # breaks = seq(0,1,by=0.2)
    labels = scales::comma
    , limits = c(0,NA)
    , expand = ggplot2::expansion(mult = c(0,0.1))
  ) +
  ggplot2::facet_grid(
    cols = dplyr::vars(scenario_desc_fact)
    , rows = dplyr::vars(name)
    # , ncol = 2
    , axes = "all"
    , scales = "free_y"
    # , switch = "y"
  ) +
  ggplot2::labs(
    x = ""
    , y = ""
    , fill = ""
    , subtitle = "comparison of height and diameter distributions\ncut versus leave trees"
  ) +
  ggplot2::theme_light() +
  ggplot2::theme(
    legend.position = "none"
    , strip.text = ggplot2::element_text(size = 11, color = "black", face = "bold")
    , axis.text.x = ggplot2::element_text(
      color = "black", face = "bold"
      # size = 9, angle = 45, hjust = 1, vjust = 1
    )
    # , axis.text.x = ggplot2::element_text(size = 12)
    # , axis.text.y = ggplot2::element_blank()
    # , axis.ticks.y = ggplot2::element_blank()
  )
```

table that

```{r}
trees2ico_ans_trees_df %>% 
  sf::st_drop_geometry() %>% 
  dplyr::select(scenario_desc_fact, cut_leave, tree_height_m, dbh_cm) %>% 
  dplyr::group_by(
    scenario_desc_fact, cut_leave
  ) %>% 
  dplyr::summarise(
    dplyr::across(
      dplyr::where(is.numeric)
      , .fns = list(
        mean = ~mean(.x,na.rm=T)
        , sd = ~sd(.x,na.rm=T)
        # , q10 = ~quantile(.x,na.rm=T,probs=0.1)
        # , q50 = ~quantile(.x,na.rm=T,probs=0.5)
        # , q90 = ~quantile(.x,na.rm=T,probs=0.9)
        # , min = ~min(.x,na.rm=T)
        # , max = ~max(.x,na.rm=T)
        , range = ~paste0(
          scales::comma(min(.x,na.rm=T), accuracy = 0.1)
          ,"-"
          , scales::comma(max(.x,na.rm=T), accuracy = 0.1)
        )
      )
    )
    , n = dplyr::n()
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::select(-c(n)) %>% 
  dplyr::mutate(
    dplyr::across(
      dplyr::where(is.numeric)
      , ~scales::comma(.x,accuracy=0.1)
    )
  ) %>% 
  # ncol()
  kableExtra::kbl(
    caption = "comparison of height and diameter distributions <br />cut versus leave trees"
    , col.names = c(
      "scenario", "cut/leave"
      , rep(c("Mean","Std Dev","Range"), times = 2)
    )
    , escape = F
  ) %>% 
  kableExtra::kable_styling(font_size = 12) %>% 
  kableExtra::add_header_above(
    c(
      " "=2
      , "Height (m)" = 3
      , "DBH (cm)" = 3
    )
    , escape = F
  ) %>% 
  kableExtra::column_spec(seq(2,8,by=3), border_right = TRUE, include_thead = TRUE) %>% 
  kableExtra::collapse_rows(columns = 1, valign = "top")  
```


