## Manitou EF ICO Prescription

let's use the spatial tree list inventory for implementing the Individuals, Clumps, and Openings (ICO) method defined by Churchill et al. ([2016](https://scholarworks.umt.edu/ico/3)) to generate a cut-retention prescription. The `trees2ico` [package and workflow](https://github.com/georgewoolsey/trees2ico/tree/main) determines which trees to 'cut' or 'keep' based on stand-level targets for basal area and specific clump size proportions.

load the `trees2ico` [package](https://github.com/georgewoolsey/trees2ico/tree/main) which was developed specifically for this use-case

```{r, eval=FALSE}
## remotes helps us get packages hosted on github
install.packages("remotes")
## get trees2ico
remotes::install_github(repo = "georgewoolsey/trees2ico", upgrade = F)
```

### Define Targets

in order to define our basal area and tree clump group proportion targets, we need to know the current conditions. let's use some `trees2ico` helper functions to get the current conditions as we follow xxx

first, we'll define our distance in meters to separate tree clump groupings. See [Churchill et al. (2016)](https://scholarworks.umt.edu/ico/3) Figure 4 (p.10) and [Matonis and Binkley (2018)](https://scholar.google.com/scholar?cluster=14613232085713826018&hl=en&as_sdt=0,6) who "calculated coverage of mosaic-meadows (percentage of stand > 6 m from overstory trees)" (p. 124)

```{r}
my_clump_dist_m <- 6
ico_clumps_temp <- trees2ico:::get_ico_clumps(treetops_sf, clump_dist_m = my_clump_dist_m) 
# huh?
names(ico_clumps_temp)
```

this gave us the tree list with clumps defined using a maximum distance between trees for determining tree clumps of `r scales::comma(my_clump_dist_m,accuracy=.1)` m which is based on the approximate crown radius of a typical, mature overstory tree. [Churchill et al. (2016)](https://scholarworks.umt.edu/ico/3) provide instructions for implementing the clump identification (Plotkin et al. 2002) in ArcGIS:

>Use the Buffer tool (in the Proximity toolset within the Analysis toolbox) to create a buffer of distance $d/2$, one half the inter-tree distance, around each point. This quantity $d/2$ is meant to approximate the crown radius of a "typical" overstory tree. Set the Dissolve Type option to ALL, which dissolves overlapping buffers, creating a reduced set of spatially non-overlapping polygons stored as a multipart polygon feature...[Sanchez Meador et al. (2011)](https://scholar.google.com/scholar?cluster=15721451396848616769&oi=gsb&hl=en&as_sdt=0,6) provide some useful examples of how clump attributes can be summarized...The method described here can be modified to use measured or modeled crown radii for each tree in place of $d/2$
(p.36)

with respect to clump size groupings, [Churchill et al. (2016)](https://scholarworks.umt.edu/ico/3) note that:

>Proportions for clump sizes should be lumped into four or five bins for operational simplicity. We use 4 or 5 bins (Fig 5): individual trees, small clumps (2-4 trees), medium clumps (5-9 trees), and large clumps (10-20+ trees). Note that when instructed to leave a large clump (e.g. 10-20 trees), marking crews often have difficulty leaving the upper end of the size range (e.g. an 18, 19, or 20 tree clump). Thus adding a fifth bin for “super clumps” may be necessary (e.g. 15-20 trees or 20-25+ trees), especially if the upper size range of clumps is desired. (p. 12-13)

`trees2ico` implements these group sizes by default

```{r}
ico_clumps_temp$tree_list %>% 
  dplyr::pull(clump_n_trees_grp) %>% 
  # get factor levels in case there aren't any represented in our tree list
  levels()
```

use another `trees2ico` helper function to see the current distribution of trees by clump size in our data as well as the BA

```{r}
agg_stand_clumps_temp <- 
  ico_clumps_temp %>% 
  purrr::pluck("tree_list") %>%
  # get rid of columns we use as param
  dplyr::select( -dplyr::any_of(c(
    "hey_xxxxxxxxxx"
    , "stand_area_ha"
  ))) %>%
  trees2ico:::agg_stand_clumps(
    stand_area_ha = stand_sf$stand_area_ha
  )
```

current (pre-treatment) distribution of trees by clump size

```{r}
agg_stand_clumps_temp$props %>%
  dplyr::mutate(
    count=scales::comma(count,accuracy=1)
    , prop=scales::percent(prop,accuracy=0.1)
  ) %>% 
  kableExtra::kbl(
    caption = "current (pre-treatment) group size tree distribution"
    , col.names = c(
      "group size","# trees","% trees"
    )
  ) %>% 
  kableExtra::kable_styling()
```

what about BA?

```{r}
print(paste(
  "current stand BA:"
  , scales::comma(agg_stand_clumps_temp$stats$ba_m2_ha, accuracy = 0.1)
  , "m2/ha"
))
```

starting from the current clump size proportions and basal area, we'll set targets for our ICO prescription based on historical reference conditions with relevant research for our study area and forest type from:

* [Churchill et al. (2013)](https://doi.org/10.1016/j.foreco.2012.11.007)
* [Brown et al. (2015)](https://doi.org/10.1139/cjfr-2014-0387)
* [Ziegler et al. (2017)](https://doi.org/10.1016/j.foreco.2016.12.002)

define 3 target scenarios

```{r}
ico_targets_df <- 
  levels(ico_clumps_temp$tree_list$clump_n_trees_grp) %>% 
    dplyr::as_tibble() %>% 
    dplyr::rename(clump_n_trees_grp = 1) %>% 
    dplyr::mutate(
      clump_n_trees_grp = ordered(clump_n_trees_grp, levels =  levels(ico_clumps_temp$tree_list$clump_n_trees_grp))
    ) %>% 
    tidyr::uncount(3, .id = "scenario_id") %>% 
    dplyr::arrange(scenario_id, clump_n_trees_grp) %>% 
  dplyr::bind_cols(
    # scenarios
    dplyr::bind_rows(
      dplyr::tibble(
        scenario_desc = "Low Clumping (Churchill et al. 2013)"
        , target_props = c(0.42, 0.37, 0.16, 0.05, 0.0)
        , target_ba_m2_ha = 15
      )
      , dplyr::tibble(
          scenario_desc = "Moderate Clumping (Ziegler et al. 2017)"
          , target_props = c(
            median(c(0.27,0.25,0.28,0.36,0.51,0.14,0.58))
            , median(c(0.38,0.36,0.37,0.35,0.24,0.28,0.42))
            , median(c(0.25,0.30,0.25,0.16,0.12,0.27,0.0))
            , 0.0 # median(c(0.11,0.10,0.06,0.08,0.12,0.15,0.0))
            , 0.0
          )
          , target_ba_m2_ha = median(c(14.5,19.7,10.6,9.5,9.0,17.6,7.5))
        ) %>% 
        dplyr::mutate(
          target_props = ifelse(
            dplyr::row_number() == 4
            , (1 - sum(target_props))
            , target_props
          )
        )
      , dplyr::tibble(
        scenario_desc = "High Clumping (Churchill et al. 2016 [Malheur])"
        , target_props = c(0.15, 0.25, 0.25, 0.1, 0.25)
        , target_ba_m2_ha = 21
      )
    )
  ) %>% 
  # set a constant BA target?
    # Brown et al. (2015): 25% q = 14.8, mean = 16.7; 75% q = 21.5
  dplyr::mutate(
    target_ba_m2_ha = 17
    , scenario_desc = scenario_desc %>% 
      stringr::str_extract("^.*?(?=\\()") %>% 
      stringr::str_squish() %>% 
      forcats::fct_inorder()
    # , huh = as.numeric(scenario_desc)
  )
# dplyr::glimpse(ico_targets_df)
```



```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```
