# Appendix

use raw point clouds to generate a spatial tree list inventory for use implementing the Individuals, Clumps, and Openings (ICO) method defined by Churchill et al. ([2016](https://scholarworks.umt.edu/ico/3)) to generate a cut-retention prescription. The `trees2ico` [package and workflow](https://github.com/georgewoolsey/trees2ico/tree/main) determines which trees to 'cut' or 'keep' based on stand-level targets for basal area and specific clump size proportions.

## Multi-Stand ICO Thinning

what if we have a huge tree list across multiple stands?

The `trees2ico` [package and workflow](https://github.com/georgewoolsey/trees2ico/tree/main) determines which trees to 'cut' or 'keep' based on stand-level targets for basal area and specific clump size proportions. It operates specifically on a single stand and it depends on the analyst to pre-define the trees available for cutting within the stand. The algorithm simply uses the tree list it is provided and the given targets and attempts to minimize the divergence of the residual conditions to the targets based on the user-provided required stand area and the optional DBH limit.

load the `trees2ico` [package](https://github.com/georgewoolsey/trees2ico/tree/main) which was developed specifically for this use-case

```{r, eval=FALSE}
## remotes helps us get packages hosted on github
install.packages("remotes")
## get trees2ico
remotes::install_github(repo = "georgewoolsey/trees2ico", upgrade = F)
```

Load the standard libraries we use to do work

```{r, warning=FALSE, message=FALSE}
# bread-and-butter
library(tidyverse) # the tidyverse
library(viridis) # viridis colors
library(harrypotter) # hp colors
library(RColorBrewer) # brewer colors
library(scales) # work with number and plot scales
library(latex2exp)

# visualization
library(mapview) # interactive html maps
library(kableExtra) # tables
library(patchwork) # combine plots

# spatial analysis
library(terra) # raster
library(sf) # simple features
library(lidR) # lidar data
library(cloud2trees) # the cloud2trees
```

```{r, include=F, warning=F, message=F}
#########################
#   ```{r,eval=F
#########################

# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results = 'hide'
  , fig.width = 10.5
  , fig.height = 7
)
# option to put satellite imagery as base layer of mapview maps
  mapview::mapviewOptions(
    homebutton = FALSE
    # , basemaps = c("Esri.WorldImagery","OpenStreetMap")
    , basemaps = c("OpenStreetMap", "Esri.WorldImagery")
  )
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

## Data

let's check out the data we might have if we processed a point cloud covering a broad area with multiple management units/stands within the footprint of the point cloud data used to derive the `cloud2trees` spatial tree list

where is all of this data?

```{r}
input_dir <- "../data/Stand_Thinning_Function"
out_dir <- "../data/Stand_Thinning_Function/outputs"
# we can define a tree height threshold to work with only "overstory" trees
my_min_tree_height_m <- 2
# we also need to define the distance threshold used to identify clump groups and openings
my_clump_dist_m <- 6
```

read in a large tree list data as points (tree tops)

```{r}
cloud2trees_tree_tops <-
  list.files(
    input_dir
    , pattern = "final_detected_tree_tops.*\\.gpkg$"
    , full.names = T
  ) %>% 
  normalizePath() %>% 
  purrr::map(
    \(x)
    sf::st_read(dsn = x, quiet = T)
  ) %>% 
  dplyr::bind_rows() %>% 
  # get rid of things we'll create...just in case
  dplyr::select( -dplyr::any_of(c(
    "hey_xxxxxxxxxx"
    , "stand_area_ha"
    , "stand_id"
  ))) %>% 
  # work with overstory
  dplyr::filter(dplyr::coalesce(tree_height_m,0) >= my_min_tree_height_m) %>% 
  dplyr::slice_sample(prop = 0.2)
# huh?
dplyr::glimpse(cloud2trees_tree_tops)
```

say we have a single spatial file (`.gpkg`, `.shp`, etc.) with multiple stand polygon boundaries which we identify with a column in the data named `stand_id`

```{r}
# Read in stand polygons
stand_boundaries <- sf::st_read(file.path(input_dir,"Extent_West_Units.gpkg"), quiet = T) %>% 
  dplyr::rename_with(tolower) %>% 
  dplyr::rename_with(stringr::str_squish) %>% 
  dplyr::rename_with(make.names) %>% 
  dplyr::rename_with(~stringr::str_replace_all(.x, "\\.{2,}", ".")) %>% 
  dplyr::rename_with(~stringr::str_remove(.x, "\\.$")) %>% 
  dplyr::rename_with(~stringr::str_replace_all(.x, "\\.", "_")) %>% 
  # transform to tree list proj
  sf::st_transform(sf::st_crs(cloud2trees_tree_tops)) %>% 
  sf::st_set_geometry("geometry") %>%
  dplyr::mutate(stand_id = poly_name) %>% 
  # if stands are in multiple records, union to be safe
  dplyr::group_by(stand_id) %>%
  dplyr::summarise(
    geometry = sf::st_union(geometry)
    , .groups = "drop"
  ) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(
    stand_area_ha = sf::st_area(.) %>% as.numeric() %>% `/`(10000)
  )
# what?
stand_boundaries %>% 
  dplyr::select(stand_id, stand_area_ha) %>% 
  dplyr::glimpse()
```

where in the world are we and what data is this?

```{r}
# colors
pal_stands <- 
  unique(stand_boundaries$stand_id) %>% 
  length() %>% 
  viridis::inferno(begin = 0.1, end = 0.9) %>% 
  setNames(
    unique(stand_boundaries$stand_id) %>%
    as.character()
  )
# scales::show_col(pal_stands)
# map
mapview::mapview(
  cloud2trees_tree_tops %>% 
    dplyr::select(tree_height_m) %>% 
    dplyr::slice_sample(n = 2222)
  , zcol = "tree_height_m"
  , col.regions = viridis::mako(n=100, direction = -1)
  , layer.name = "sample trees"
  , cex = 3
) +
mapview::mapview(
  stand_boundaries %>% dplyr::select(stand_id)
  , zcol = "stand_id"
  , color = pal_stands
  , layer.name = "stand bounds"
  , alpha.regions = 0
  , lwd = 2
)

```

## Get ICO Prescriptions{#prescriptions}

for each stand, the `trees2ico` package requires a stand average basal area (BA) target in square meters per hectare as well as targets for the proportion of trees in different size clump groups

it is straightforward that target basal area is a single number. To be clear, users will need to define targets for each of the following clump sizes with example proportions that should sum to 100% listed in the table below

| Group Size      | Target Proportion of Trees |
|-----------------|----------------------------|
| Individual      | 28%                      |
| 2–4 trees       | 36%                      |
| 5–9 trees       | 25%                      |
| 10–15 trees     | 11%                      |
| >15 trees       | 0%                       |

A user-friendly way to define ICO prescriptions is to define the targets for each stand in a spreadsheet where each row contains the stand targets

we'll load in an example target spreadsheet file

```{r}
prescriptions_sheet <- readr::read_csv(file.path(input_dir, "xxxStand_Thinning_Prescriptions.csv"), show_col_types = F, progress = F) %>% 
  dplyr::rename_with(tolower) %>% 
  dplyr::rename_with(stringr::str_squish)  
# print the table as it appears in the spreadsheet
prescriptions_sheet %>% 
  kableExtra::kbl(
    caption = "example stand prescription spreadsheet format"
  ) %>% 
  kableExtra::kable_styling(font_size = 11.5)
```

we'll assume that the prescriptions entered in the spreadsheet contain no errors (i.e. no incorrect or missing column names, columns are all in the same order as displayed here, proportions sum to 1, one record per stand, each target has a matching stand in the stand polygon data, etc.), so won't have a data quality check here.

we can automatically convert from imperial units to metric units and do some more data manipulation that will make it easier to programmatically iterate through the different stands

```{r}
prescriptions_cleaned <-
  prescriptions_sheet %>% 
  # we'll rely on the order of the columns in this demo instead of validating
  tidyr::pivot_longer(
    cols = -c(stand_id,target_ba_ft2_acre,dbh_cap_inch)
    , names_to = "clump_group"
    , values_to = "target_props"
    , values_drop_na = F
  ) %>% 
  dplyr::mutate(
    # convert ft^2/acre to m^2/hectare
    target_ba_m2_ha = target_ba_ft2_acre / 3.28^2 * 2.47 
    , dbh_limit_cm = 
      # convert inches to centimeters
      (dbh_cap_inch * 2.54) %>% 
      dplyr::coalesce(900)
    , target_props = dplyr::coalesce(target_props,0)
    , stand_id = as.character(stand_id)
  ) %>% 
  dplyr::select(-c(target_ba_ft2_acre,dbh_cap_inch)) %>% 
  dplyr::relocate(stand_id,target_ba_m2_ha,dbh_limit_cm)
# check out this cleaned data
prescriptions_cleaned %>% 
  dplyr::slice_head(n = 10) %>% 
  kableExtra::kbl(
    caption = "automatically cleaned example stand prescription spreadsheet format"
  ) %>% 
  kableExtra::kable_styling(font_size = 11.5)
```

`trees2ico` operates specifically on a single stand and it depends on the analyst to pre-define the trees available for cutting within the stand. The algorithm simply uses the tree list it is provided and the given targets and attempts to minimize the divergence of the residual conditions to the targets based on the user-provided required stand area and the optional DBH limit.

with only a few stands, this is simple to do manually. for example:

```{r, include = T, eval=FALSE}
# stand1_trees
stand1_trees <- sf::st_intersection(
  x = cloud2trees_tree_tops
  , y = stand_boundaries %>% dplyr::filter(stand_id=="1") %>% dplyr::select(stand_id)
)
# stand2_trees
stand2_trees <- sf::st_intersection(
  x = cloud2trees_tree_tops
  , y = stand_boundaries %>% dplyr::filter(stand_id=="2") %>% dplyr::select(stand_id)
)
```

and then we would run `trees2ico` on each stand. for example:

```{r, include = T, eval=FALSE}
# stand1_trees2ico_ans
stand1_trees2ico_ans <- trees2ico::trees2ico(
  tree_list = stand1_trees
  , target_clump_props = prescriptions_cleaned %>% 
    dplyr::filter(stand_id=="1") %>% 
    dplyr::pull(target_props)
  , target_ba_m2_ha = prescriptions_cleaned %>% 
    dplyr::filter(stand_id=="1") %>% 
    dplyr::slice_max(order_by = target_ba_m2_ha, with_ties = F, na_rm = T) %>% 
    dplyr::pull(target_ba_m2_ha)
  , dbh_limit_cm = prescriptions_cleaned %>% 
    dplyr::filter(stand_id=="1") %>% 
    dplyr::slice_max(order_by = dbh_limit_cm, with_ties = F, na_rm = T) %>% 
    dplyr::pull(dbh_limit_cm)
  , clump_dist_m = my_clump_dist_m
  , stand_area_ha = stand_boundaries %>% dplyr::filter(stand_id=="1") %>% dplyr::pull(stand_area_ha)
)
# stand2_trees2ico_ans
stand2_trees2ico_ans <- trees2ico::trees2ico(
  tree_list = stand2_trees
  , target_clump_props = prescriptions_cleaned %>% 
    dplyr::filter(stand_id=="2") %>% 
    dplyr::pull(target_props)
  , target_ba_m2_ha = prescriptions_cleaned %>% 
    dplyr::filter(stand_id=="2") %>% 
    dplyr::slice_max(order_by = target_ba_m2_ha, with_ties = F, na_rm = T) %>% 
    dplyr::pull(target_ba_m2_ha)
  , dbh_limit_cm = prescriptions_cleaned %>% 
    dplyr::filter(stand_id=="2") %>% 
    dplyr::slice_max(order_by = dbh_limit_cm, with_ties = F, na_rm = T) %>% 
    dplyr::pull(dbh_limit_cm)
  , clump_dist_m = my_clump_dist_m
  , stand_area_ha = stand_boundaries %>% dplyr::filter(stand_id=="2") %>% dplyr::pull(stand_area_ha)
)
```

let's combine these steps into a single function which filters the trees for each stand and then implements the `trees2ico` algorithm using the stand-specific targets (technically we could make this multiple functions, but we'll just do it quickly all in one)

## Function to loop over stands

minimum required inputs:

* a huge `cloud2trees` tree list across multiple stands
* a polygon spatial data set with one row per stand and the `stand_id` column
* a [prescription data set](#prescriptions) with one row per stand and the `stand_id` column

we'll assume that all the data is correct and not perform and quality checks right now for the demo

```{r}
trees2ico_stands <- function(
  stand_polys
  , unfiltered_tree_list
  , prescriptions
  , clump_dist_m = 6
  , outdir = tempdir()
  , get_just_trees = T
) {
  # map over the stands
  ans <- unique(prescriptions$stand_id) %>% 
    purrr::set_names() %>% 
    purrr::map(function(this_stand_id){
      message( paste0("trees2ico for stand_id: ", this_stand_id, " .... started at", Sys.time()) )
      ################################################
      # could do this in-line but i'll be explicit to filter the datas
      ################################################
      # prescriptions
      this_prescriptions <- prescriptions %>% dplyr::filter(stand_id==this_stand_id)
      # stand_polys
      this_stand_polys <- stand_polys %>% dplyr::filter(stand_id==this_stand_id)
      
      ################################################
      # trees in stand
      ################################################
      this_stand_trees <- sf::st_intersection(
        x = unfiltered_tree_list
        , y = this_stand_polys %>% dplyr::select(stand_id)
      ) %>% 
      # get rid of things we'll create...just in case
      dplyr::select( -dplyr::any_of(c(
        "hey_xxxxxxxxxx"
        , "stand_area_ha"
        , "stand_id"
      )))
      
      ################################################
      # trees2ico
      ################################################
      stand_trees2ico_ans <- trees2ico::trees2ico(
        tree_list = this_stand_trees
        , target_clump_props = this_prescriptions$target_props # assumes correct order
        , target_ba_m2_ha = this_prescriptions$target_ba_m2_ha[1] # just 1
        , dbh_limit_cm = this_prescriptions$dbh_limit_cm[1] # just 1
        , clump_dist_m = clump_dist_m
        , stand_area_ha = this_stand_polys$stand_area_ha[1] # just 1
      )
      
      ################################################
      # write
      ################################################
      # fnms
      fnm_tree_list <- file.path(outdir, paste0("tree_list_stand_id_",this_stand_id,".gpkg") )
      fnm_pre_cut_clumps <- file.path(outdir, paste0("pre_cut_clumps_stand_id_",this_stand_id,".gpkg") )
      fnm_post_cut_clumps <- file.path(outdir, paste0("post_cut_clumps_stand_id_",this_stand_id,".gpkg") )
      # write
      stand_trees2ico_ans$tree_list %>% sf::st_write( fnm_tree_list, append = F, quiet = T )
      stand_trees2ico_ans$pre_cut_clumps %>% sf::st_write( fnm_pre_cut_clumps, append = F, quiet = T )
      stand_trees2ico_ans$post_cut_clumps %>% sf::st_write( fnm_post_cut_clumps, append = F, quiet = T )
      
      return(stand_trees2ico_ans)
    })
  
  ################################################
  # return just the tree list 
  ################################################
  if(get_just_trees){
    just_trees <- 
      purrr::map(ans, ~ purrr::pluck(.x, "tree_list")) %>% 
      purrr::list_rbind(names_to = "stand_id") %>% 
      dplyr::relocate("stand_id")
  
    return(just_trees)
  }else{
    return(ans)
  }
  
}
```

## `trees2ico_stands`

test `trees2ico_stands` out

```{r, results=F}
if(!dir.exists(out_dir)) dir.create(out_dir)
# if(length(list.files(out_dir))==0){} # we only need to do this once for the demo
trees2ico_stands_ans <- trees2ico_stands(
  stand_polys = stand_boundaries
  , unfiltered_tree_list = cloud2trees_tree_tops
  , prescriptions = prescriptions_cleaned
  , clump_dist_m = my_clump_dist_m
  , outdir = out_dir
  , get_just_trees = T
)


# add on stand ha
trees2ico_stands_ans <- trees2ico_stands_ans %>% 
  dplyr::left_join(
    stand_boundaries %>% 
      sf::st_drop_geometry() %>% 
      dplyr::select(stand_id,stand_area_ha)
  )
```

check out the residual BA and clump size proportions

```{r, include = F, eval = F}
# count props
levels(trees2ico_stands_ans$clump_n_trees_grp) %>% 
  dplyr::as_tibble() %>% 
  dplyr::rename(clump_n_trees_grp = 1) %>% 
  dplyr::mutate(
    clump_n_trees_grp = ordered(clump_n_trees_grp, levels =  levels(trees2ico_stands_ans$clump_n_trees_grp))
  ) %>% 
  tidyr::crossing(stand_boundaries %>% sf::st_drop_geometry() %>% dplyr::select(stand_id)) %>% 
  tidyr::crossing(dplyr::tibble(status=c("pre-treatment","post-treatment"))) %>% 
  
  
  dplyr::bind_rows(
    # get pre-treatment
      trees2ico_stands_ans %>% 
      sf::st_drop_geometry() %>%
      dplyr::group_by(stand_id,stand_area_ha,pre_clump_n_trees_grp) %>% 
      dplyr::summarise(
        basal_area_m2_per_ha=sum(basal_area_m2)
        , trees = dplyr::n()
        , .groups = "drop"
      ) %>% 
      dplyr::ungroup() %>% 
      dplyr::rename(clump_n_trees_grp=pre_clump_n_trees_grp) %>% 
      dplyr::group_by(stand_id) %>% 
      dplyr::mutate(
        trees = dplyr::coalesce(trees,0)
        , pct_trees = trees/sum(trees,na.rm = T)
        , basal_area_m2_per_ha = sum(basal_area_m2_per_ha,na.rm = T)/stand_area_ha
      ) %>% 
      dplyr::ungroup() %>% 
      dplyr::mutate(status = "pre-treatment")
    # get post-treatment
    ,
      trees2ico_stands_ans %>% 
      sf::st_drop_geometry() %>%
      dplyr::group_by(stand_id,stand_area_ha,post_clump_n_trees_grp) %>% 
      dplyr::summarise(
        basal_area_m2_per_ha=sum(basal_area_m2)
        , trees = dplyr::n()
        , .groups = "drop"
      ) %>% 
      dplyr::ungroup() %>% 
      dplyr::rename(clump_n_trees_grp=post_clump_n_trees_grp) %>% 
      dplyr::group_by(stand_id) %>% 
      dplyr::mutate(
        trees = dplyr::coalesce(trees,0)
        , pct_trees = trees/sum(trees,na.rm = T)
        , basal_area_m2_per_ha = sum(basal_area_m2_per_ha,na.rm = T)/stand_area_ha
      ) %>% 
      dplyr::ungroup() %>% 
      dplyr::mutate(status = "post-treatment")
  ) 
# table
pre_post_temp %>% 
  dplyr::select(-c(trees)) %>% 
  dplyr::mutate(
    pct_trees = scales::percent(pct_trees,accuracy=1)
    , basal_area_m2_per_ha = scales::comma(basal_area_m2_per_ha,accuracy=.1)
  ) %>% 
  tidyr::pivot_wider(names_from = clump_n_trees_grp, values_from = pct_trees) %>% 
  dplyr::rename(scenario=stand_id) %>% 
  dplyr::arrange(scenario,desc(status)) %>% 
  dplyr::relocate(scenario,status) %>% 
  dplyr::rename(
    `BA m<sup>2</sup> ha<sup>-1</sup>` = basal_area_m2_per_ha
  ) %>% 
  kableExtra::kbl(
    caption = "ICO Scenarios: pre- and post-treatment proportion of trees in different sized clumps"
    , escape = F
    # , digits = 2
  ) %>% 
  kableExtra::kable_styling() %>% 
  kableExtra::collapse_rows(columns = 1, valign = "top")
```

